### 2017-02-02 Manuscript/SW_early_life_telomeres.Rmd
Edits

#####Study species and sampling
The Seychelles warbler is a small (~15 g), insectivorous passerine bird with a mean life expectancy of 5.5 years at fledging [@Hammers2013]. The population of *ca*. 320 birds on Cousin Island (04'20'S, 55'40'E) has been intensively studied since 1986 [@Richardson2003b; @Spurgin2014a]. This species' main breeding season runs from June-September, when the breeding females on each of the *ca*. 115 territories lay one or, rarely, two or three eggs [@Komdeur1991]. As a result of this low reproductive output, combined with higher mortality in first-year birds [39% in first-year birds versus 16% in adults; @Brouwer2006], cohort sizes in the Seychelles warbler are typically small (< 50).

Individuals are usually ringed (with an individually numbered metal ring and unique combination of colour rings) in their first year of life, and so are of known age. We aged all birds using information on eye colour [@Komdeur1991a] and previous captures [@Richardson2003b]. As Seychelles warblers are non-migratory endemics naturally confined to the island [@Komdeur2004a], a biannual census of birds on Cousin during each breeding season gives accurate measures of local density, social status (e.g. breedier, helper, non-helper) and individual survival [@VandeCrommenacker2011; @Barrett2013a]. Full details of monitoring methods can be found in Brouwer *et al.* [-@Brouwer2012].

Seychelles warblers are highly territorial and all territories were mapped during each main breeding season using detailed observational data of foraging and territorial defence behaviour, and surveyed for territory quality [@Richardson2003b]. Territory quality is calculated based on territory size, foliage cover and insect abundance [@Komdeur1992]. Where territory quality estimates were not available for a specific year we used the average value for that territory across years [@Hammers2013]. Cousin is subject to intra- and inter-annual variation in rainfall and food availability, and such island-wide temporal variation may override the effects of individual territory quality. As an estimate of seasonal variation in food availability, we calculated an index of the number of insects across the entire island during each main breeding season (referred to hereafter as 'insect abundance'). This index is calculated as the mean number of insects found per unit leaf area over all surveys carried out on the island in a main breeding season.

Each time a bird is caught on Cousin body mass and tarsus length are measured (to the nearest 0.1g and 0.1mm, respectively). A blood sample (*ca* 25 $\mu$l) is taken from each bird captured via brachial venipuncture, and stored at room temperature in 1 ml of absolute ethanol in a 1.5 ml screw-cap microfuge tube.

#####Molecular methods
For each sample, genomic DNA was extracted from a ~2 mm^2^ flake of preserved blood using the DNeasy Blood and Tissue Kit (Qiagen), following the manufacturer's protocol, with the modification of overnight lysis at 37^o^C and a final DNA elution volume of 80 $\mu$l. Sex was determined using the PCR-based method outlined by Griffiths *et al.* [-@Griffiths1998]. Prior to telomere analysis, DNA concentration and purity were quantified using a NanoDrop 8000 Spectrophotometer (ThermoScientific), and the following thresholds were applied before samples were included for further analysis: i) DNA concentration must be at least 15 ng $\mu$l^-1^ (based on a mean of three measurements), ii) the 260/280 absorbance ratio has to be between 1.8 and 2 for acceptable DNA purity and, iii) the 260/230 absorbance ratio must be higher than 1.8. DNA integrity was further validated by visualization with ethidium bromide after electrophoresis on a 1.2% agarose gel, and all samples with evidence of DNA degradation were re-extracted or excluded. All DNA extractions that passed the above criteria were diluted to 3.3 ng $\mu$l^-1^ before telomere measurement. We measured relative telomere length (RTL) for all samples using a quantitative PCR (qPCR) assay of telomeres and a GAPDH control gene, following Bebbington *et al.* [-@Bebbington2016].

For a subset of birds we had longitudinal data, with one or more additional samples taken at different ages (n = `r length(unique(c(dd3$BloodID,dd3$BloodID1)))` measurements from `r length(unique(dd3$BirdID))` birds). For these individuals we calculated the within-individual change in RTL by subtracting RTL at time point *t* from RTL at time point *t* + 1 (hereafter $\Delta$RTL, n = `r nrow(dd3)` measurements). Negative values of $\Delta$RTL reflect decreases in telomere length with age, while positive values increases.

#####Statistical analyses
We performed all statistical analyses using R version 3.2.2 [@RDevelopmentCoreTeam2011]. RTL was square root transformed to improve linear model fits, and we assessed repeatability of RTL using the rptR package. We explored the cross-sectional relationship between RTL and age among cohorts using linear mixed models (LMMs) carried out in the lme4 package [@Bates2014]. Following a similar approach to Fairlie *et al.* [-@Fairlie2016], we compared a selection of models fitting different relationships between RTL and age. We created models where the relationship between RTL and age was linear, quadratic, log-linear, and where age was fitted as a factor. For each age term, we fitted additional models including birth year (cohort) as a factor, and an interactions between cohort age. All fitted models are included in Table 1. For random effects we included individual ID, and catch year. Models were compared using AIC with correction for finite sample size (AICc).

@@

When examining the distribution of longitudinal telomere changes we observed increases in telomere length with age, so we used repeat measurements to test whether these increases could be explained by measurement error. We calculated the change in RTL between pairs of repeat measurements within samples (hereafter $\Delta$RTL~*sample*~; N = `r nrow(subset(ddL,Group == 'Within samples'))` measurements from `r length(unique(subset(ddL,Group == 'Within samples')$BirdID))` samples) in exactly the same way as for across samples (hereafter $\Delta$RTL~*individual*~), using completely separate reactions run on separate plates. To test whether greater changes in RTL were observed among individuals compared to among repeat samples, we compared the variance in $\Delta$RTL~*sample*~ and $\Delta$RTL~*individual*~ using a Levene's test. Then, to separately test whether the extent of telomere increases and decreases within individuals was greater than expected by measurement error, we split $\Delta$RTL measurements into groups in which RTL decreased ($\Delta$RTL < 0) and increased $\Delta$RTL ($\Delta$RTL > 0), and tested whether $\Delta$RTL~*inidividual*~ values were significantly different from $\Delta$RTL~*sample*~ values, using Wilcoxon tests.

We then used LMMs to test how variation in environmental and social conditions influenced telomere length within cohorts. We created a full model with RTL as a response variable, alongside the following explanatory variables: log age (based on the RTL and age analysis; see results), tarsus length, body mass, sex, insect abundance, territory quality, island-wide population density (an annual measure estimated from the summer breeding census), territory group size, and the number of helping subordinate birds present in the territory. The random effects structure was informed by the analysis of telomere dynamics and age (see results): we included individual ID, cohort ID, and a random slope of log age among cohorts. We report model estimates and confidence intervals for all effects included in the full model. We also calculated marginal R^2^ and conditional R^2^ [incorporating fixed and random effects, respectively; @Johnson2014] to assess the explanatory power of these models. As a complementary approach, we also performed model averaging, using the MuMIn package in R [@Barton2012]. Model selection was performed using the full model described above and a top model set defined, containing all models with AICc $\leq$ 6 compared to the best supported model [@Burnham2011]. We report model-averaged coefficients, confidence intervals and 'relative importance', which reflects the relative weights of each predictor variable across the top model set.

@@

###Results
We measured telomere lengths using a total of `r length(unique(dd$BloodID))` unique samples from juvenile and adult Seychelles warblers from `r length(table(dd$Cohort))` cohorts born between `r min(dd$LayYear)` and `r max(dd$LayYear)`. Efficiencies (mean $\pm$ s.d.) for our telomere and GAPDH reactions were (`r round2(mean(dd$EffTelomere))` $\pm$ `r round2(sd(dd$EffTelomere))`) and (`r round2(mean(dd$EffGAPDH))` $\pm$ `r round2(sd(dd$EffGAPDH))`) respectively. Inter-plate repeatability of RTL, based on `r nrow(subset(ddL,Group == 'Within samples'))` samples measured at least twice, was `r round2(rptall$R)` (CI = `r round2(rptall$CI_emp[1])`, `r round2(rptall$CI_emp[2])`).

#####Telomere dynamics and age amnong cohorts
We first tested how RTL was related to age among cohorts using a model selection approach. The top model contained a loglinear relationship between RTL and age, as well as a log age x cohort interaction (Table 1A). The second-best model contained log age and cohort ID, with no interaction term, although this model was substantially poorer in terms of model fit ($\Delta$AICc > 10). All other models fitted the data much less well ($\Delta$AICc > 35; Table 1). The log-linear relationship between RTL and age could be seen clearly in the raw data; RTL decreased with age (estimate = `r round3(fixef(RTL_LogAge)['LogAge'],lessthan = F)`, CIs = `r paste(round3(CIwald(RTL_LogAge)['LogAge',],lessthan = F),collapse = ', ')`), with the greatest decrease occurring in the first year of life (Fig. 1A). There was substantial variation in RTL among cohorts, with no obvious trend over time (Fig. 1B). There was a negative relationship between RTL and age in 21 of the `r length(table(dd$Cohort))` cohorts, but the slope of the relationship between log age and RTL varied substantially among cohorts (Fig.1C). In the one year in which RTL increased with age (2013), 17 of the 18 birds were fledglings or subadults, suggesting that the observed pattern was an artifact of the sampling in this season rather than a real relationship.

@@

There was positive correlation between telomere measurements taken at different time points (Fig. 2A), but this was very weak (marginal R^2^ = `r round2(r.squaredGLMM(RTL_RTL1)['R2m'])`), and not significant (estimate = `r round3(fixef(RTL_RTL1)['RTL'],lessthan = F)`, CIs = `r paste(round3(CIwald(RTL_RTL1)['RTL',],lessthan = F),collapse = ', ')`). Although both cross-sectional and longitudinal data indicated a general trend of telomere shortening with age, we found that RTL within individuals increased in `r round((nrow(subset(dd3,DeltaRTL>0))/nrow(dd3))*100,0)`% of our `r nrow(dd3)` $\Delta$RTL~*individual*~ measurements (Fig. 2A). To test whether increases in telomere length in our dataset could be explained by measurement error, we compared variance in telomere length among repeat measurements of samples to the variance observed among different samples of the same individual. We found significantly higher variance in telomere length over individual lifetimes compared to among sample replicates (Levene's test: F = `r round2(leveneTest(DeltaRTL~Group,data=ddL)['F value'][1,1])`; P `r round3(leveneTest(DeltaRTL~Group,data=ddL)['Pr(>F)'][1,1])`; Fig. 2B). Splitting the longitudinal data into instances of decreasing (i.e. $\Delta$RTL < 0) and increasing (i.e. $\Delta$RTL > 0) telomere length revealed that not only did we observe significantly greater decrease in RTL within individuals compared to within samples (Wilcoxon test: P `r round3(wilcox.test(DeltaRTL~Group,data=subset(ddL,DeltaRTL<0))$p.value)`), but also a significantly greater increase (P `r round3(wilcox.test(DeltaRTL~Group,data=subset(ddL,DeltaRTL>0))$p.value)`; Fig. 2B).

>>>>

#####Study species and sampling
The Seychelles warbler is a small (~15 g), insectivorous passerine bird with a mean life expectancy of 5.5 years at fledging [@Hammers2013]. The population of *ca*. 320 adult birds on Cousin Island (04'20'S, 55'40'E) has been intensively studied since 1986 [@Komdeur1992; @Richardson2003b; @Spurgin2014a]. ). This species' main breeding season runs from June-September (though a small number of territories do sometimes try to breed between January-March) when the breeding females on many of the ca. 110 territories will attempt to breed, laying one or, rarely, two or three eggs [@Komdeur1991]. Breeding attempts are often unsuccessful, and as a result of this low reproductive output, and higher mortality in first-year birds [39% in first-year birds versus 16% in adults; @Brouwer2006], cohort sizes in the Seychelles warbler are typically small (< 50). The 22 birth year cohorts used in this study cover 1993 to 2014 – the time period during which our data and sampling is most complete.

The majority of individuals are ringed (with an individually numbered metal ring and unique combination of colour rings) within the first year of life, and so are of known age. We aged all birds using information on eye colour at first capture [@Komdeur1991a] and previous capture history [@Richardson2003b]. As Seychelles warblers are non-migratory endemics naturally confined to the island [@Komdeur2004a], an extensive biannual census of birds on Cousin during each breeding season gives accurate measures of local density, social status (e.g. breeder, helper, non-helper) and individual survival [@VandeCrommenacker2011; @Barrett2013a]. Full details of monitoring methods can be found in Brouwer *et al.* [-@Brouwer2012].

Seychelles warblers are highly territorial and all territories were mapped during each main breeding season using detailed observational data of foraging and territorial defence behaviour [@Richardson2003b]. Territory quality is calculated based on territory size, foliage cover and insect abundance [@Komdeur1992]. Where territory quality estimates were not available for a specific year we used the average value for that territory across years [@Hammers2013]. Cousin is subject to considerable intra- and inter-annual variation in rainfall and, consequently, insect availability. Such island-wide temporal variation may override the effects of variation in individual territory quality across the island. As an estimate of seasonal variation in food availability, we calculated an index of the abundance of insects across the entire island during each main breeding season (referred to hereafter as 'insect abundance'). This index is calculated as the mean number of insects found per unit leaf area over all monthly surveys carried out on the island in a main breeding season.

Each time a bird is caught on Cousin a range of morphometric measurements are taken, including body mass and tarsus length (to the nearest 0.1g and 0.1mm, respectively). A blood sample (ca 25 l) is taken via brachial venipuncture, and stored at room temperature in 1 ml of absolute ethanol in a 1.5 ml screw-cap microfuge tube.

#####Molecular methods
For each sample, genomic DNA was extracted from a ~2 mm^2^ flake of preserved blood using the DNeasy Blood and Tissue Kit (Qiagen), following the manufacturer's protocol, with the modification of overnight lysis at 37^o^C and a final DNA elution volume of 80 $\mu$l. Sex was determined using the PCR-based method outlined by Griffiths *et al.* [-@Griffiths1998]. Prior to telomere analysis, DNA concentration and purity were quantified using a NanoDrop 8000 Spectrophotometer (ThermoScientific). The following thresholds were applied before samples were included for further analysis: i) DNA concentration must be at least 15 ng $\mu$l^-1^ (based on a mean of three measurements), ii) the 260/280 absorbance ratio has to be between 1.8 and 2 for acceptable DNA purity and, iii) the 260/230 absorbance ratio must be higher than 1.8. DNA integrity was further validated by visualization with ethidium bromide after electrophoresis on a 1.2% agarose gel, and all samples with evidence of DNA degradation were re-extracted or excluded. All DNA extractions that passed the above criteria were diluted to 3.3 ng $\mu$l^-1^ before telomere measurement. We measured relative telomere length (RTL) for all samples using a quantitative PCR (qPCR) assay of telomeres and a GAPDH control gene, following Bebbington *et al.* [-@Bebbington2016]. Based on the distribution of observed cq values, we excluded outlier samples with extremely large cq values (cq values > 25 and 26 were excluded for the telomere and GAPDH reactions, respectively), which were assumed to be failed reactions.

For a large subset of birds we had longitudinal data, with two or more samples taken at different ages (n = `r length(unique(c(dd3$BloodID,dd3$BloodID1)))` measurements from `r length(unique(dd3$BirdID))` birds). For these individuals we calculated the within-individual change in RTL by subtracting RTL at time point *t* from RTL at time point *t* + 1 (hereafter $\Delta$RTL, n = `r nrow(dd3)` measurements). Negative values of RTL reflect decreases in telomere length with age, while positive values reflect increases.

#####Statistical analyses
We performed all statistical analyses using R version 3.2.2 [@RDevelopmentCoreTeam2011]. RTL was square root transformed to improve linear model fits, and we assessed repeatability of RTL using the rptR package.

We explored the cross-sectional relationship between RTL and age among cohorts using linear mixed models (LMMs) carried out in the lme4 package [@Bates2014]. Following a similar approach to Fairlie *et al.* [-@Fairlie2016], we compared a selection of models fitting different relationships between RTL and age. We created models where the relationship between RTL and age was linear, quadratic, log-linear, and where age was fitted as a factor. For each age term, we fitted additional models including birth year (cohort) as a factor, and an interaction between cohort age. All fitted models are included in Table 1. Note that we do not carry out full model selection or model averaging here, as our aim was to compare a set of specifically defined models. For random effects we included individual ID, and catch year. Models were compared using AIC with correction for finite sample size [AICc; @Hurvich1989].

@@

When examining the distribution of longitudinal telomere changes we observed some increases in telomere length with age in individuals. We then used repeat measurements to test whether these increases could be explained by measurement error. We calculated the change in RTL between pairs of repeat measurements within the same samples (hereafter $\Delta$RTL~*sample*~; N = `r nrow(subset(ddL,Group == 'Within samples'))` measurements from `r length(unique(subset(ddL,Group == 'Within samples')$BirdID))` samples) in exactly the same way as for across samples (hereafter $\Delta$RTL~*individual*~), using completely separate reactions run on separate plates. To test whether greater changes in RTL were observed among individuals compared to among repeat samples, we compared the variance in $\Delta$RTL~*sample*~ and $\Delta$RTL~*individual*~ using a Levene's test. Then, to separately test whether the extent of telomere increases and decreases within individuals was greater than expected by measurement error, we split $\Delta$RTL measurements into groups in which RTL decreased ($\Delta$RTL < 0) and increased $\Delta$RTL ($\Delta$RTL > 0), and tested whether $\Delta$RTL~*inidividual*~ values were significantly different from $\Delta$RTL~*sample*~ values, using Wilcoxon tests.

We then used LMMs to test how variation in environmental and social conditions influenced telomere length within cohorts. We created a full model with RTL as a response variable, alongside the following explanatory variables: log age (based on the RTL and age analysis; see results), tarsus length, body mass, sex, insect abundance, territory quality, island-wide population density (an annual measure estimated from the summer breeding census), territory group size, and the number of helping subordinate birds present in the territory. The random effects structure was informed by the analysis of telomere dynamics and age (see results): we included individual ID, cohort ID, and a random slope of log age among cohorts (to allow the effect of age on RTL to vary among cohorts). We report model estimates and confidence intervals for all effects included in the full model. We also calculated marginal R^2^ and conditional R^2^ [incorporating fixed and random effects, respectively; @Johnson2014] to assess the explanatory power of these models. As a complementary approach, we also performed model averaging, using the MuMIn package in R [@Barton2012]. Model selection was performed using the full model described above and a top model set defined, containing all models with AICc $\leq$ 6 compared to the best supported model [@Burnham2011]. We report model-averaged coefficients, confidence intervals and 'relative importance', which reflects the relative weights of each predictor variable across the top model set.

@@

###Results
We measured telomere lengths using a total of `r length(unique(dd$BloodID))` unique samples from juvenile and adult Seychelles warblers from `r length(table(dd$Cohort))` cohorts born between `r min(dd$LayYear)` and `r max(dd$LayYear)`. Efficiencies (mean $\pm$ s.d.) for our telomere and GAPDH reactions were (`r round2(mean(dd$EffTelomere))` $\pm$ `r round2(sd(dd$EffTelomere))`) and (`r round2(mean(dd$EffGAPDH))` $\pm$ `r round2(sd(dd$EffGAPDH))`) respectively. Inter-plate repeatability of RTL, based on `r nrow(subset(ddL,Group == 'Within samples'))` samples measured at least twice at different timepoints, was `r round2(rptall$R)` (CI = `r round2(rptall$CI_emp[1])`, `r round2(rptall$CI_emp[2])`).

#####Telomere dynamics and age among cohorts
We first tested how RTL was related to age among cohorts using a model selection approach. The top model contained a loglinear relationship between RTL and age, as well as a log age x cohort interaction (Table 1A). The second-best model contained log age and cohort ID, with no interaction term, although this model was substantially poorer in terms of model fit ($\Delta$AICc > 10). All other models fitted the data much less well ($\Delta$AICc > 35; Table 1). The log-linear relationship between RTL and age could be seen clearly in the raw data; RTL decreased with age (estimate = `r round3(fixef(RTL_LogAge)['LogAge'],lessthan = F)`, CIs = `r paste(round3(CIwald(RTL_LogAge)['LogAge',],lessthan = F),collapse = ', ')`), with the greatest decrease occurring in the first year of life (Fig. 1A). There was substantial variation in RTL among cohorts, with no obvious trend over time (Fig. 1B). There was a negative relationship between RTL and age in 21 of the `r length(table(dd$Cohort))` cohorts, but the slope of the relationship between log age and RTL varied substantially among cohorts (Fig.1C). In the one year in which RTL increased with age (2013), 17 of the 18 birds sampled were fledglings or subadults, suggesting that the observed pattern was an artifact of the sampling in this season (i.e. a lack of variation in age among sampled birds) rather than a real relationship.

@@

There was positive correlation between RTL measured from different samples taken at different time points during an individual’s life (Fig. 2A), but this was very weak (marginal R^2^ = `r round2(r.squaredGLMM(RTL_RTL1)['R2m'])`), and not significant (estimate = `r round3(fixef(RTL_RTL1)['RTL'],lessthan = F)`, CIs = `r paste(round3(CIwald(RTL_RTL1)['RTL',],lessthan = F),collapse = ', ')`). Although both cross-sectional and longitudinal data indicated a general trend of telomere shortening with age, we found that RTL - measured across two samples taken from the same individuals over time - increased with age in `r round((nrow(subset(dd3,DeltaRTL>0))/nrow(dd3))*100,0)`% of our `r nrow(dd3)` $\Delta$RTL~*individual*~ measurements (Fig. 2A). To test whether increases in telomere length in our dataset could be explained by measurement error, we compared variance in telomere length among repeat measurements of the same samples to the variance observed among different samples of the same individual. We found significantly higher variance in telomere length over individual lifetimes compared to among sample replicates (Levene's test: F = `r round2(leveneTest(DeltaRTL~Group,data=ddL)['F value'][1,1])`; P `r round3(leveneTest(DeltaRTL~Group,data=ddL)['Pr(>F)'][1,1])`; Fig. 2B). Splitting the longitudinal data into instances of decreasing (i.e. $\Delta$RTL < 0) and increasing (i.e. $\Delta$RTL > 0) telomere length revealed that not only did we observe significantly greater decrease in RTL within individuals compared to within samples (Wilcoxon test: P `r round3(wilcox.test(DeltaRTL~Group,data=subset(ddL,DeltaRTL<0))$p.value)`), but also a significantly greater increase (P `r round3(wilcox.test(DeltaRTL~Group,data=subset(ddL,DeltaRTL>0))$p.value)`; Fig. 2B).
